Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-20 paper/paper.tex
applied Ewa's comments

Scientific workflows, usually represented as directed acyclic graphs (DAG), are
the important class of applications that have been studied in the context of
resource management and scheduling on grid and utility computing systems.
However, large computational applications are often composed of several 
inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ by
input data, number of tasks and individual tasks sizes. A good example of
scientific workflow ensemble comes from the CyberShake
application~\cite{Callaghan11}, which calculates seismic hazards for a given
geographical region such as California. In order to produce a hazard map, the
hazard curves need to be computed for a set of geographical sites, while each
site requires running a single complex scientific workflow. Workflows for each
site may differ not only by their parameters, but also by priority; e.g. some of
the points on the map correspond to urban areas or strategic locations such as
power plants, whereas others may be located in less populated regions. Another
example of workflow ensembles comes from astronomical applications, such as
Montage~\cite{Deelman08} or search for Earth-like planets from Kepler project
data~\cite{vockler11}, where multiple, different workflows are used to
process data for different areas of the sky.

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost, on-demand computing resources~\cite{Ostermann10,Keahey09}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give users more flexibility to creating a controlled and managed computing environment
with the ability to adjust resource capacity according to the dynamically changing
computing demands of the application, often called auto-scaling. Giving the users 
more control, however, also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as workflow makespan or resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) cloud providers 
are usually not free~\cite{Durkee10}.

In this paper, to get insight into these resource management challenges for
scientific workflows on clouds, we address a new important problem of maximizing
the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation is to answer the fundamental question of a
researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble due to the constraints. As methodology in the study of the
proposed algorithms we use simulation techniques. We have developed cloud workflow simulator based on
CloudSim~\cite{Calheiros11}, which models the infrastructure and the application. The
algorithms are subject to evaluation on a set of scientific workflow ensembles with a broad range of
budget and deadline parameters.

The paper is organized as follows: after analysis of related work in
Section~\ref{sec:related}, Section~\ref{sec:problem} gives the details about the
infrastructure and application model we are targeting in this paper.
Subsequently, in Section~\ref{sec:algorithms} we describe the dynamic and
static algorithms we developed. Section~\ref{sec:performance} presents test
scenarios, metrics and results of simulation studies of these algorithms.
Finally, general conslusions, lessons learned and future work is outlined in
Section~\ref{sec:conclusions}.

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
Typical infrastructure-specific metric is system load, whereas
application-specific metrics include response time and queue length. It is
possible to set thresholds and limits to tune the behavior of these autoscaling
systems, but no support for complex applications is provided.

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniqies. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of scheduling heuristics that cannot be applied to independent
tasks.

@@

We should refer to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan10,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
trated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes tham less useful in practical applications. Our work can be also
regarded as extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In this model, virtual machine (VM) instances are requested on-demand. The VMs are billed by the hour, with partial hours being rounded up. Although there are multiple VM types with different amounts of CPU, memory, disk space, and I/O, for this paper we focus on a single VM type because we assume that for most applications there will typically be one VM type with the best price/performance ratio for the application.

@@

This paper focuses on scheduling and provisioning for workflow ensembles. An ensemble consists of several related workflows that a user wishes to run. Each workflow in an ensemble is given a priority that indicates how important the workflow is to the user.

@@

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spds-gantt}
 \caption{Example schedule generated using SPDS algorithm. Tasks are colored by workflow. }
\label{fig:spds-example}
\end{figure}

@@

The static provisioning approach of SPDS may not be optimal when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing 
is that the number of resources can be changed to meet the demands of an application.
In policy-based provisioning (or auto-scaling) systems, metrics such as resource 
utilization and queue length are monitored to estimate application demand. The DPDS 
algorithm uses this approach to determine when to provision and de-provision VMs. The 
algorithm periodically computes resource utilization as the percentage of idle VMs, and
adjusts the number of VMs if the utilization is above or below given thresholds.
Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate. This process
is described in Algorithm~\ref{alg:prov}.

@@

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated on line \ref{l:nT2} are chosen to increase the resource utilization
to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the number of running VMs is greater than the product of
$N_{VM}$ and an autoscaling factor, $v_{max}$.

@@

The DPDS algorithm does not use any information about the structure of the workflows
in the ensemble when scheduling tasks: it looks only at the priorities of the ready 
tasks when deciding which task to schedule next. It does not consider whether a lower 
priority task belongs to a workflow that it will never be able to complete given the 
current budget and deadline. As a result, DPDS may execute lower priority tasks just 
to keep VMs busy that will end up delaying higher priority tasks, making it less likely
that higher priority workflows will be able to finish.

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends 
DPDS by introducing a workflow admission procedure. The admission procedure is
invoked whenever WA-DPDS sees the first task of a new workflow at the head of the 
priority queue (i.e. when no other tasks from the workflow have been scheduled yet). 
The admission procedure---shown in Algorithm~\ref{alg:wa-dpds}---estimates if there is 
enough budget remaining to admit the new workflow; if there is not, then the 
workflow is rejected and its tasks are removed from the queue. This algorithm 
compares the current cost (consumed budget) and remaining budget, taking
into account the cost of currently running VMs, and the cost of workflows 
that have already been admitted. It adds a small safety margin avoid going over
the budget.

@@

This admission procedure relies only on the total estimated resource consumption and
compares it to the budget remaining. We found that this estimation is useful not
only to prevent low-priority workflows from delaying high-priority ones, but
also to reject large and costly workflows that would overrun the budget and
admit smaller workflows that can efficiently utilize idle resources in ensembles 
containing workflows of non-uniform sizes. It would also be possible to extend this 
admission procedure to check other constraints, such as whether the estimated 
critical path of the new workflow exceeds the time remaining until the deadline.

@@

The approach used by SPSS is to plan each workflow in the ensemble in priority 
order, rejecting any workflows that cannot be completed by the deadline or that 
cause the cost of the plan to exceed the budget. Once the plan is complete, then 
the VMs are provisioned and tasks are executed according to the schedules given
by the plan.

>>>>

Scientific workflows, usually represented as directed acyclic graphs (DAG), are
the important class of applications that have been studied in the context of
resource management and task scheduling on grid and utility computing systems.
However, large computational applications are often composed of several 
inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ by
input data, number of tasks and individual task sizes. A good example of
scientific workflow ensemble comes from the CyberShake
application~\cite{Callaghan11}, which calculates seismic hazards for a given
geographical region such as California. In order to produce a hazard map, the
hazard curves need to be computed for a set of geographical sites, while each
site requires running a single complex scientific workflow. Workflows for each
site may differ not only in their parameters, but also in their priority; e.g. some of
the points on the map correspond to urban areas or strategic locations such as
power plants, whereas others may be located in less populated regions. Another
example of workflow ensembles comes from astronomical applications, such as
Montage~\cite{Deelman08} or search for exo-solar planets from Kepler project
data~\cite{vockler11}, where multiple, different workflows are used to
process data for different areas of the sky.

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost, on-demand computing resources~\cite{Ostermann10,Keahey09}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give users more flexibility to creating a controlled and managed computing environment.
Clouds provide the ability to adjust resource capacity according to the dynamically changing
computing demands of the application, often called auto-scaling. However, giving users 
more control over the resources also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as the workflow makespan or the resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) cloud providers 
usually have monetary costs associated with them~\cite{Durkee10}.

In this paper, we aim to gain insight into resource management challenges when executing 
scientific workflow ensembles on clouds. In particular we address a new important problem of maximizing
the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation is to answer the fundamental question of a
researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble due to the constraints. We evaluate the algorithms via simulation. 
We have developed cloud workflow simulator based on
CloudSim~\cite{Calheiros11}, which models the infrastructure and the application. The
algorithms are subject to evaluation on a set of scientific workflow ensembles with a broad range of
budget and deadline parameters.

The paper is organized as follows: after analysis of related work in
Section~\ref{sec:related}, Section~\ref{sec:problem} gives the details about the
infrastructure and application model we are targeting in this paper.
In Section~\ref{sec:algorithms} we describe the dynamic and
static algorithms we developed. Section~\ref{sec:performance} presents test
scenarios, metrics and results of simulation studies of these algorithms.
Finally, general conclusions, lessons learned and future work are outlined in
Section~\ref{sec:conclusions}.

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
A typical infrastructure-specific metric is system load, whereas
application-specific metrics include response time and length of task or
request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniqies. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of workflow-aware scheduling heuristics that cannot be applied to
independent tasks.

@@

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan10,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in practice. Our work can be also
regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In
this model, virtual machine (VM) instances are requested on-demand. The VMs are
billed by the hour, with partial hours being rounded up. Although there are
multiple VM types with different amounts of CPU, memory, disk space, and I/O,
for this paper we focus on a single VM type because we assume that for most
applications there will typically be one VM type with the best price/performance
ratio for the application~\cite{Juve2009}. We also assume that a submitted task
has exclusive access to a VM instance and that there is no preemption.

@@

This paper focuses on scheduling and provisioning for workflow ensembles. A
workflow ensemble consists of several related workflows that a user wishes to
run. Each workflow in an ensemble is given a priority that indicates how
important the workflow is to the user.

@@

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spds-gantt}
 \caption{Example schedule generated using SPDS algorithm. The colors indicate how various tasks from different workflows are scheduled onto Virtual Machines. }
\label{fig:spds-example}
\end{figure}

@@

The static provisioning approach of SPDS may not be optimal when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing 
is that the number of resources can be changed to meet the demands of an application.
In policy-based provisioning (or auto-scaling) systems, metrics such as resource 
utilization or task queue length are monitored to estimate application demand.
The DPDS algorithm uses this approach to determine when to provision and de-provision VMs. The 
algorithm periodically computes resource utilization as the percentage of idle VMs, and
adjusts the number of VMs if the utilization is above or below given thresholds.
Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate. This process
is described in Algorithm~\ref{alg:prov}.

@@

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated on line \ref{l:nT2} are chosen to increase the resource utilization
to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the number of running VMs is greater than the product of
$N_{VM}$ and an autoscaling factor, $v_{max}$ that has to be set manually.

@@

The DPDS algorithm does not use any information about the structure of the workflows
in the ensemble when scheduling tasks: it looks only at the priorities of the ready 
tasks when deciding which task to schedule next. It does not consider whether a lower 
priority task belongs to a workflow that will never be able to complete given the 
current budget and deadline. As a result, DPDS may execute lower priority tasks just 
to keep VMs busy and that will end up delaying higher priority tasks, making it less likely
that higher priority workflows will be able to finish.

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends 
DPDS by introducing a workflow admission procedure. The admission procedure is
invoked whenever WA-DPDS sees the first task of a new workflow at the head of the 
priority queue (i.e. when no other tasks from the workflow have been scheduled yet). 
The admission procedure---shown in Algorithm~\ref{alg:wa-dpds}---estimates whether there is 
enough budget remaining to admit the new workflow; if there is not, then the 
workflow is rejected and its tasks are removed from the queue. This algorithm 
compares the current cost (consumed budget) and remaining budget, taking
into account the cost of currently running VMs, and the cost of workflows 
that have already been admitted. It adds a small safety margin to  avoid going over
the budget.

@@

This admission procedure relies only on the total estimated resource consumption and
compares it to the remaining budget. We found that this estimation is useful not
only to prevent low-priority workflows from delaying high-priority ones, but
also to reject large and costly workflows that would overrun the budget and
admit smaller workflows that can efficiently utilize idle resources in ensembles 
containing workflows of non-uniform sizes. It would also be possible to extend this 
admission procedure to check other constraints, such as whether the estimated 
critical path of the new workflow exceeds the time remaining until the deadline.

@@

The approach used by SPSS is to plan each workflow in the ensemble in priority 
order, rejecting any workflows that cannot be completed by the deadline or that 
cause the cost of the plan to exceed the budget. Once the plan is complete, then 
the VMs are provisioned and tasks are executed according to the schedules given
by the plan. However, in this paper we only consider a deterministic environment and task times.
